---
title: "Data Analysis - Clustering"
author: "Varun Bansal"
output: pdf_document
date: "2023-02-15"
---


## Setting the work directory. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE)
options(scipen=9)
knitr::opts_chunk$set(fig.width=9, fig.height=7,
                      fig.path='Figs/', echo = TRUE)
#This sets the working directory
knitr::opts_knit$set(root.dir = 'C:/Users/varun/OneDrive/Desktop/R/datasets')

rm(list = ls())

```


## Loading and attaching all the necessary packages.

```{r}
#Load packages

if(!require(tinytex)){
  install.packages("tinytex")
  library("tinytex")
}

if (!require(lattice)) {
  install.packages("lattice")
  library(lattice)
}

if (!require(gridExtra)) {
  install.packages("gridExtra")
  library(gridExtra)
}

```

# Data Transformation

```{r}

# Reading File
df <- read.csv("Expense_Summary.txt", header = TRUE, sep = ",")
head(df, 5)

```

## Renaming all variables with my initials appended

```{r}

# Appending initials to all variables in the data frame 

df_VB <- df
colnames(df_VB) <- paste(colnames(df_VB), "VB", sep = "_")
head(df_VB, 5)

```

## Standardizing the variables

Before we do standardization, lets look at the summary of data.

```{r}
summary(df_VB)
```
There doesn't seem to be much of outliers.  
Lets do the Shapiro-Wilk test to check for normality.

```{r}
shapiro.test(df_VB$Food_VB)
shapiro.test(df_VB$Enter_VB)
shapiro.test(df_VB$Edu_VB)
shapiro.test(df_VB$Trans_VB)
shapiro.test(df_VB$Work_VB)
shapiro.test(df_VB$House_VB)
shapiro.test(df_VB$Oth_VB)
```

Here in all the test, p-value is less than 0.05.
So data is not normally distributed.

Also our data has a bounded range, so we will use min-max standardization function.  
  
  
Using min-max standardization function
```{r}

# min-max standardization function

norm01 <- function(x) {
  return ((x - min(x)) / (max(x) - min(x)))
}

```


```{r}

# performing min-max standardization function on all the variables

df_VB$Food_Norm01_VB <- norm01(df_VB$Food_VB)
df_VB$Enter_Norm01_VB <- norm01(df_VB$Enter_VB)
df_VB$Edu_Norm01_VB <- norm01(df_VB$Edu_VB)
df_VB$Trans_Norm01_VB <- norm01(df_VB$Trans_VB)
df_VB$Work_Norm01_VB <- norm01(df_VB$Work_VB)
df_VB$House_Norm01_VB <- norm01(df_VB$House_VB)
df_VB$Oth_Norm01_VB <- norm01(df_VB$Oth_VB)

```
```{r}
head(df_VB,3)
```



#  Descriptive Data Analysis
## Graphical summaries of the data 

Exploring the Data using boxplots.
```{r}

par(mfrow=c(2,2))

for (i in 1:7) {
  if (is.numeric(df_VB[,i])) {
      boxplot(df_VB[i], main=names(df_VB)[i],
              horizontal=TRUE, pch=10,
              col= 27)
  }
}

```

```{r}
par(mfrow=c(1,1))
```

The box plot suggests that there are not many outliners in the dataset.

Now lets density plot and see if data is normally distributed.

```{r}

dp1 <- densityplot( ~ df_VB$Food_VB, pch=10, main='Food_VB')
dp2 <- densityplot( ~ df_VB$Enter_VB, pch=10, main='Enter_VB')
dp3 <- densityplot( ~ df_VB$Edu_VB, pch=10, main='Edu_VB')
dp4 <- densityplot( ~ df_VB$Trans_VB, pch=10, main='Trans_VB')
dp5 <- densityplot( ~ df_VB$Work_VB, pch=10, main='Work_VB')
dp6 <- densityplot( ~ df_VB$House_VB, pch=10, main='House_VB')
dp7 <- densityplot( ~ df_VB$Oth_VB, pch=10, main='Oth_VB')

# Display the plots in a grid
grid.arrange(dp1, dp2, dp3, dp4, dp5, dp6, dp7, ncol = 2)

```


This also suggests data is not normally distributed.


# 3. Clustering  
# Using the K-Means procedure clustering

Setting Up for Clusters
```{r}

# Creating Variable for Elbow Chart
# Creating 2 to 7 Clusters

maxk <- 7   # max number of k

nk <- c(2:maxk) # list of numbers 2 to 7

wss <- rep(0,maxk-1)  # empty list having 7 zeros
```

  
  
Creating Clusters
```{r}

# Setting Clusters 2 to 7

for(k in 2:7){
  
  ClstrIncome_VB <- kmeans(df_VB[,c(8,13)], iter.max=10, centers=k, nstart=10)
  
  cat("****************  k = ",k, " *******************\n\n")
  cat("Cluster size: ", ClstrIncome_VB$size, "\n\n")
  cat("Cluster Centers: \n")
  print(ClstrIncome_VB$centers)
  cat("\nRatio of between-cluster variance to total variance ", 
      ClstrIncome_VB$betweenss/ClstrIncome_VB$totss)
  cat("\n\n---------------------------------------------------------------\n\n")
  
  df_with_k <- paste("df_VB", k, sep="_")
  df_VB$cluster <- factor(ClstrIncome_VB$cluster)   # Adding Cluster tags to variables
  assign(df_with_k, df_VB)
  
  centers <- paste("centers", k, sep="_")
  
  # the data frame and assign it to the name
  assign(centers, data.frame(cluster=factor(1:k), ClstrIncome_VB$centers))  
  wss[k-1] <- ClstrIncome_VB$tot.withinss
}
```

\

## 3. Creating the WSS plots

Plotting 'Elbow' chart
```{r}

plot(2:maxk, wss,
     type="b", 
     pch = 15, 
     col="blue",
     frame = FALSE,
     main="Elbow Chart for Clusters",
     xlab="Values of K",
     ylab="Total Within-Clusters SS",
     ylim=c(0,max(wss)))
```

Looking at the elbow chart, there seems to be a bend at 4.  
So we choose the value of k as 4.

#  Evaluation of Clusters
## Plotting the clusters

We have choosen k=4.  
Plotting the clusters for k=3, k=4, k=5
```{r}

# K=3

plot(df_VB_3$House_Norm01_VB, df_VB_3$Food_Norm01_VB,
     col=df_VB_3$cluster, pch=as.numeric(df_VB_3$cluster),
     main = "k = 3",
     xlab= "House",
     ylab = "Food")

points(centers_3$House_Norm01_VB, centers_3$Food_Norm01_VB,
    col=centers_3$cluster, 
    pch=as.numeric(centers_3$cluster),
    cex=3, lwd=3)

# K=4

plot(df_VB_4$House_Norm01_VB, df_VB_4$Food_Norm01_VB,
     col=df_VB_4$cluster, pch=as.numeric(df_VB_4$cluster),
     main = "k = 4",
     xlab= "House",
     ylab = "Food")

points(centers_4$House_Norm01_VB, centers_4$Food_Norm01_VB,
    col=centers_4$cluster, 
    pch=as.numeric(centers_4$cluster),
    cex=3, lwd=3)


# K=5

plot(df_VB_5$House_Norm01_VB, df_VB_5$Food_Norm01_VB,
     col=df_VB_5$cluster, pch=as.numeric(df_VB_5$cluster),
     main = "k = 5",
     xlab= "House",
     ylab = "Food")

points(centers_5$House_Norm01_VB, centers_5$Food_Norm01_VB,
    col=centers_5$cluster,
    pch=as.numeric(centers_5$cluster),
    cex=3, lwd=3)


```

\

Looking at the WSS plot and the charts, at k = 3, clusters look the best and well segregated.  


## Summarizing the Clusters
```{r}

# Creating summary report

SumClusters_VB <- aggregate(
  cbind(Food_VB, Enter_VB, Edu_VB, Trans_VB, Work_VB, House_VB, Oth_VB) ~ cluster, 
  df_VB_3, 
  FUN = mean)

SumClusters_VB

```


## Suitable descriptive names for each cluster.

For cluster 1: High on transport and housing, negligible on education and entertainment.

For cluster 2: High on housing, low on entertainment and education.

For cluster 3: High on education, negligible on work.


## Uses for this clustering scheme.

There can be many uses of this clustering scheme. Some of them are- 

This scheme may come handy in making business strategies. For example, if a company primarily sells products related to housing and transportation, they may want to expand their offerings to appeal to customers in cluster 1.
Similarly, if a company is developing a new product related to education, they may want to focus on customers in cluster 3, who are likely to spend more in this area.

This clustering scheme may also help government in there policy-making. They can identify areas where public spending should be prioritized. For example, if cluster 3 represents a large portion of the population, policymakers may want to invest more resources into education to meet the needs of this group. 
This may also help the government to analyse why people are not willing to spend on education in Cluster 1 and 


