---
title: "Exploratory Data Analysis"
author: "Varun Bansal"
output: pdf_document
date: "2023-01-18"
---

## Setting the work directory. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(warning=FALSE)
options(scipen=9)
knitr::opts_chunk$set(fig.width=9, fig.height=7,
                      fig.path='Figs/', echo = TRUE)
#This sets the working directory
knitr::opts_knit$set(root.dir = 'C:/Users/varun/OneDrive/Desktop/R/datasets')

```


## Loading and attaching all the necessary packages.

```{r}
#Load packages

if(!require(tinytex)){
  install.packages("tinytex")
  }
library("tinytex")

if (!require(dplyr)) {
  install.packages("dplyr")
  library(dplyr)
}

```



## Basic Manipulation

## Reading the text file and changing it to a data frame

```{r}
df <- read.csv("Data_Centers.txt", header = TRUE, sep = ",")
```

## Append the initials to all variables in the data frame 

```{r}
df_VB <- df
colnames(df_VB) <- paste(colnames(df_VB), "VB", sep = "_")
head(df_VB, 5)
```


## Changing each character variable to a factor variable

```{r}
# Changing each character variable to factor variable
df_VB <- as.data.frame(unclass(df_VB), stringsAsFactors = TRUE)

str(df_VB)
```

## Checking dimensions of the dataset

```{r}
dim(df_VB)
```
The dataset is of dimension 90000 x 6 (90000 rows and 6 columns)


## Summarizing Data

## Means and Standard Deviations for Server Message Blocks Received.

```{r}
mean_smbr_VB <- mean(df_VB$SMBR_VB)
paste("Mean of Server Message Blocks Received:", mean_smbr_VB)

sd_smbr_VB <- sd(df_VB$SMBR_VB)
paste("Standard Deviation of Server Message Blocks Received:", sd_smbr_VB)
```


## Calculating the coefficient of variation.

```{r}
cv_smbr_VB <- round((sd_smbr_VB / mean_smbr_VB) * 100, 3)
paste("Coefficient of variation of Server Message Blocks Received:", cv_smbr_VB)
```

## Calculating the mean and standard deviation for Server Message Blocks Transmitted.

```{r}
mean_smbt_VB <- mean(df_VB$SMBT_VB)
paste("Mean of Server Message Blocks Transmitted:", mean_smbt_VB)

sd_smbt_VB <- sd(df_VB$SMBT_VB)
paste("Standard Deviation of Server Message Blocks Transmitted:", sd_smbt_VB)
```


## Calculating the coefficient of variation (rounded to 3 decimal places).

```{r}
cv_smbt_VB <- round((sd_smbt_VB / mean_smbt_VB) * 100, 3)
paste("Coefficient of variation of Server Message Blocks Transmitted:", cv_smbt_VB)
```


## Looking for which variable has more variation (SMBT or SMBR) 

```{r}
max(cv_smbr_VB,cv_smbt_VB)
```
SMBT has more variation, as coefficient of variation of SMBT is double the coefficient of variation of SMBR.


## Organizing Data

## Summary Table

## Table showing the average Server Message Blocks Transmitted by Manufacturer.

```{r}
avg_smbt_df_VB <- aggregate(df_VB$SMBT_VB, by=list(df_VB$Manufacturer_VB), 
                            FUN=function(x) round(mean(x), 2))
colnames(avg_smbt_df_VB) <- c("Manufacturer", "AverageSMBT")
avg_smbt_df_VB
```


## Looking for which Manufacturer’s Servers have, on average, transmitted the most server message blocks.

```{r}
head( avg_smbt_df_VB[order(avg_smbt_df_VB$AverageSMBT, decreasing = TRUE), ], 1)
```
Lled has transmitted the most server message blocks on an average.



##  Cross Tabulation
##  Table counting all Servers by Data Centre.

```{r}
table_servers_by_DC_VB <- table(df_VB$Server_VB, df_VB$DC_VB)
count_servers_by_DC_VB <-margin.table(table_servers_by_DC_VB, 2)
count_servers_by_DC_VB
```


## Changing the table to show the percentage of each Server in each Data Centre.

```{r}
percentage_servers_by_DC_VB <- round((prop.table(count_servers_by_DC_VB))*100,3)
percentage_servers_by_DC_VB
```

# Data Visualization

## Bar Plot
## Bar plot of count of Servers Models.

```{r}

# Counting no. of servers for each server.
df_table_VB <- table(df_VB$Server_VB)

# Ordering by highest count of Server Model
df_table_VB <- df_table_VB[order(df_table_VB,decreasing=TRUE)]             

# Creating the bar plot of count of Servers Models
barplot(df_table_VB,
        col=55,
        density = 70, angle = 90,
        main="Bar Plot of count of Servers Models",
        xlab = "Servers",
        ylab = "Count of Server Models",
        ylim = range(df_table_VB)
        )

```


## Histogram
## Histogram of Server Message Blocks Transmitted.

```{r}

hist(df_VB$SMBT_VB,
     col=2,
     breaks=10,
     xlab="Range of SMBT",
     ylab="Count of SMBT",
     main="Histogram of Server Message Blocks Transmitted")
```



## Box plot
## Horizontal box plot of number of Connections Made.


```{r}
boxplot(df_VB$Conn_VB, 
        main="Box Plot of number of Connections Made",
        xlab="Connections Made",
        col=19,
        horizontal=TRUE,
        pch=3)

```


## Scatter Plot
## Scatter plot comparing Server Message Blocks Transmitted and Connections Made.

```{r}

plot(df_VB$SMBT_VB ~ df_VB$Conn_VB,
     data=df_VB,
     col=2,
     pch=1,
     main="Scatter plot comparing SMBT and Connections Made",
     xlab="Server Message Blocks Transmitted",
     ylab="Connections Made")

```
  
Server Message Blocks Transmitted and Connections Made have a linear relationship. It looks like it has a  moderate positive correlation.




